Rusty Variation

### 2018-06-15 doc/main.tex
Final version of SRC submission

\section*{Problem \& Motivation}
In concurrent programming, processes commonly exchange information by sending messages on shared channels. However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.

@@

\section*{Background \& Related Work}
\Textcite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by \textcite{gay2009}. Their implementation guarantees that programs correctly implement specified communication protocols with some caveats. LAST depends upon linear types, meaning that sessions have to be used exactly once. Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped. The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}. In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \texttt{\#[must\_use]} annotation. However, in general, this is not something the compiler can guarantee. Furthermore, LAST ignores the possibility of communication failures, such as network errors, which cannot be detected at compile-time, yet can cause a channel to be dropped.

@@

\section*{Contribution: Rusty Variation}
Our contribution, Rusty Variation, is an implementation of EGV in Rust. We claim that it is a faithful implementation, which preserves much of EGVs metatheory. However, we depart from it in two ways:

@@

\subsection*{Sessions and duality}
Any type built using the primitive types below is a valid session:
\begin{lstlisting}
pub struct End;
pub struct Send<T, S: Session>{ ... }
pub struct Recv<T, S: Session>{ ... }
\end{lstlisting}
Session \emph{types} are public, but their members are not. Therefore, user cannot construct their own misbehaving sessions. The continuation of a session must itself be a session. This is enforced by the \lstinline{Session} constraint. In addition to this ``kinding'' of session types, the \lstinline{Session} trait also implements duality and session generation:
\begin{lstlisting}
pub trait Session {
  type Dual: Session<Dual=Self>;
  fn new() -> (Self, Self::Dual);
}
\end{lstlisting}
Duality works as expected: \lstinline{Send<T, S>} is dual to \lstinline{Recv<T, S::Dual>}---and vice versa---and \lstinline{End} is self-dual.
The constraint \lstinline{Session<Dual=Self>} enforces that the dual of a
session is a session, and that \lstinline{S::Dual::Dual} is equal to \lstinline{S} (involutivity).

@@

\subsection*{Send, receive, and fork!}
The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. They have the following types:
\begin{lstlisting}
pub fn send<T, S: Session>(x: T, s: Send<T, S>)
       -> Result<S, Box<Error>>;
pub fn recv<T, S: Session>(s: Recv<T, S)
       -> Result<(T, S), Box<Error>>;
\end{lstlisting}
The \lstinline{fork!} macro creates a new session, and spawns a new thread. The new thread runs the first argument of \lstinline{fork!}---a code block---which receives one endpoint, and \lstinline{fork!} returns the dual endpoint. While \lstinline{fork!} is implemented as a macro, it can be considered as a function with the following type:
\begin{lstlisting}
pub fn fork!<S: Session>(p: P) -> S::Dual
      where
        P: FnOnce(S) -> Result<(), Box<Error>>;
\end{lstlisting}
Using these primitives, we can write the following program. This program forks off a process which sends a ``ping'' along the session, then waits to receive that ping and returns:
\begin{lstlisting}
let s = fork!(move |s: Send<(), End>| { send((), s) });
let ((), End) = recv(s)?;
Ok(())
\end{lstlisting}
The \lstinline{let x = M?;} construct is Rust's ``monadic bind'' notation for programs which may return errors. If \lstinline{recv(s)} fails, the \lstinline{?}-operator short-circuits, skips the rest of the statements, and returns the error.

\subsection*{Cancellation}
The \lstinline{cancel} function drops a value and cancels all sessions \emph{and} their duals referenced in that value. This cancellation propagates: any process attempting to write or write to a cancelled channel will return fail and cancel all \emph{their} open channels. This ensures that no process is left waiting for a message that will never come.
\begin{lstlisting}
pub fn cancel<T>(x: T) -> Result<(), Box<Error>>;
\end{lstlisting}
Using \lstinline{cancel}, we can implement the following program. This program forks off a process which it \emph{expects} to send a ``ping''. However, instead the forked process cancels the session. Therefore, the call to \lstinline{recv} fails, the \lstinline{?}-operator short-circuits, and the whole process returns an error:
\begin{lstlisting}
let s = fork!(move |s: Send<(), End>| { cancel(s) });
let (z, End) = recv(s)?;
Ok(())
\end{lstlisting}
The cancellation mechanism is implemented through destructors. In fact, in the current implementation it is inherited from the implementation of channels in \lstinline{std::sync::mpsc}.

@@

\subsection*{Branching with offer! and select!}
The \lstinline{offer_either}, \lstinline{select_left}, and \lstinline{select_right} functions let processes offer or make a \emph{binary} choice. They implement this by receiving or sending a value of the sum type \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session. For instance, we could write the type of a ``calculator'' server, offering negation and addition, as \lstinline{Recv<Either<NegSrv, AddSrv>, End>}.

While these functions suffice to encode choice, we also define two macros, \lstinline{offer!} and \lstinline{select!}, which generalise this functionality to \emph{all enums}---as long as each value wraps a session. This allows processes to offer and select choices between many \emph{labelled} branches, and allows us to define the type for our calculator server as:
\begin{lstlisting}
type CalcSrv = Recv<CalcOp, End>;
enum CalcOp {Neg(NegSrv), Add(AddSrv)}
\end{lstlisting}
Using these macros, we can write the following program. This program forks off a calculator server, which offers the choice between negation and addition, selects addition, sends some numbers, and then returns the result:
\begin{lstlisting}
let s = fork!(move |s: CalcSrv| { offer!(s, {
    CalcOp::Neg(s) => {
      let (x, s) = recv(s)?;
      send(-x, s)
    },
    CalcOp::Add(s) => {
      let (x, s) = recv(s)?;
      let (y, s) = recv(s)?;
      send(x + y, s)
    },})});
let s = select!(CalcOp::Add, s)?;
let s = send(4, s)?;
let s = send(5, s)?;
let (z, End) = recv(s)?;
Ok(z)
\end{lstlisting}

\subsection*{Implementation}
We use the asynchronous channels from \lstinline{std::sync::mpsc}. These are ideal, as they already implement the correct cancellation behaviour. However, they can only transfer values of a single type. Therefore, we treat them as linear channels---channels which can only be used once---and encode our communication following \textcite{dardha2017}. Each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value. Fortunately, the channels in \lstinline{std::sync::mpsc} are highly optimised for one-shot usage.

\subsection*{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Rusty Variation deadlock. The first is by looping, which we discussed earlier. The other two are both explicit, and surprisingly, \lstinline{unsafe} is not one of them:

>>>>

\section{Problem \& Motivation}
In concurrent programming, processes commonly exchange information by sending messages on shared channels. However, this communication often does not follow a specified protocol, and if it does, it the correctness of the implementation with respect to that protocol is rarely checked.

@@

\section{Background \& Related Work}
\Textcite{jespersen2015} implement session types in Rust, based on the session-typed functional language LAST by \textcite{gay2009}. Their implementation guarantees that programs correctly implement specified communication protocols with some caveats. LAST depends upon linear types, meaning that sessions have to be used exactly once. Rust only offers affine types, meaning that  values---and, particularly relevant here, sessions---can be dropped. The implementation of LAST in Rust, therefore, guarantees session fidelity under the assumption that channels are \emph{never dropped}. In some cases, the compiler can be made to issue a warning at compile-time when a channel is dropped, for instance through the \lstinline{#[must_use]} annotation. However, in general, this is not something the compiler can guarantee. Furthermore, LAST ignores the possibility of communication failures, such as network errors, which cannot be detected at compile-time, yet can cause a channel to be dropped.

@@

\section{Contribution: Rusty Variation}
Our contribution, Rusty Variation, is an implementation of EGV in Rust. We claim that it is a faithful implementation, which preserves much of EGVs metatheory. However, we depart from it in two ways:

@@

\subsection{Sessions and duality}
Any type built using the primitive types below is a valid session:
\begin{lstlisting}
pub struct End;
pub struct Send<T, S: Session>{ ... }
pub struct Recv<T, S: Session>{ ... }
\end{lstlisting}
Session \emph{types} are public, but their members are not. Therefore, user cannot construct their own misbehaving sessions. The continuation of a session must itself be a session. This is enforced by the \lstinline{Session} constraint. In addition to this ``kinding'' of session types, the \lstinline{Session} trait also implements duality and session generation:
\begin{lstlisting}
pub trait Session {
  type Dual: Session<Dual=Self>;
  fn new() -> (Self, Self::Dual);
}
\end{lstlisting}
Duality works as expected: \lstinline{Send<T, S>} is dual to \lstinline{Recv<T, S::Dual>}---and vice versa---and \lstinline{End} is self-dual.
The constraint \lstinline{Session<Dual=Self>} enforces that the dual of a
session is a session, and that \lstinline{S::Dual::Dual} is equal to \lstinline{S} (involutivity).

@@

\subsection{Send, receive, and fork!}
The \lstinline{send} and \lstinline{recv} functions send and receive values in sessions. They have the following types:
\begin{lstlisting}
pub fn send<T, S: Session>(x: T, s: Send<T, S>)
       -> Result<S, Box<Error>>;
pub fn recv<T, S: Session>(s: Recv<T, S)
       -> Result<(T, S), Box<Error>>;
\end{lstlisting}
The \lstinline{fork!} macro creates a new session, and spawns a new thread. The new thread runs the first argument of \lstinline{fork!}---a code block---which receives one endpoint, and \lstinline{fork!} returns the dual endpoint. While \lstinline{fork!} is implemented as a macro, it can be considered as a function with the following type:
\begin{lstlisting}
pub fn fork!<S: Session>(p: P) -> S::Dual
      where
        P: FnOnce(S) -> Result<(), Box<Error>>;
\end{lstlisting}
Using these primitives, we can write the following program. This program forks off a process which sends a ``ping'' along the session, then waits to receive that ping and returns:
\begin{lstlisting}
let s = fork!(move |s: Send<(), End>| { send((), s) });
let ((), End) = recv(s)?;
Ok(())
\end{lstlisting}
The \lstinline{let x = M?;} construct is Rust's ``monadic bind'' notation for programs which may return errors. If \lstinline{recv(s)} fails, the \lstinline{?}-operator short-circuits, skips the rest of the statements, and returns the error.

\subsection{Cancellation}
The \lstinline{cancel} function drops a value and cancels all sessions \emph{and} their duals referenced in that value. This cancellation propagates: any process attempting to write or write to a cancelled channel will return fail and cancel all \emph{their} open channels. This ensures that no process is left waiting for a message that will never come.
\begin{lstlisting}
pub fn cancel<T>(x: T) -> Result<(), Box<Error>>;
\end{lstlisting}
Using \lstinline{cancel}, we can implement the following program. This program forks off a process which it \emph{expects} to send a ``ping''. However, instead the forked process cancels the session. Therefore, the call to \lstinline{recv} fails, the \lstinline{?}-operator short-circuits, and the whole process returns an error:
\begin{lstlisting}
let s = fork!(move |s: Send<(), End>| { cancel(s) });
let (z, End) = recv(s)?;
Ok(())
\end{lstlisting}
The cancellation mechanism is implemented through destructors. In fact, in the current implementation it is inherited from the implementation of channels in \lstinline{std::sync::mpsc}.

@@

\subsection{Branching with offer! and select!}
The \lstinline{offer_either}, \lstinline{select_left}, and \lstinline{select_right} functions let processes offer or make a \emph{binary} choice. They implement this by receiving or sending a value of the sum type \lstinline{Either<S1, S2>}, where \lstinline{S1} and \lstinline{S2} are the possible continuations of the session. For instance, we could write the type of a ``calculator'' server, offering negation and addition, as \lstinline{Recv<Either<NegSrv, AddSrv>, End>}.

While these functions suffice to encode choice, we also define two macros, \lstinline{offer!} and \lstinline{select!}, which generalise this functionality to \emph{all enums}---as long as each value wraps a session. This allows processes to offer and select choices between many \emph{labelled} branches, and allows us to define the type for our calculator server as:
\begin{lstlisting}
type CalcSrv = Recv<CalcOp, End>;
enum CalcOp {Neg(NegSrv), Add(AddSrv)}
\end{lstlisting}
Using these macros, we can write the following program. This program forks off a calculator server, which offers the choice between negation and addition, selects addition, sends some numbers, and then returns the result:
\begin{lstlisting}
let s = fork!(move |s: CalcSrv| {
  offer!(s, {
    CalcOp::Neg(s) => {
      let (x, s) = recv(s)?;
      send(-x, s)
    },
    CalcOp::Add(s) => {
      let (x, s) = recv(s)?;
      let (y, s) = recv(s)?;
      send(x + y, s)
    },
  })});
let s = select!(CalcOp::Add, s)?;
let s = send(4, s)?;
let s = send(5, s)?;
let (z, End) = recv(s)?;
Ok(z)
\end{lstlisting}

\subsection{Implementation}
We use the asynchronous channels from \lstinline{std::sync::mpsc}. These are ideal, as they already implement the correct cancellation behaviour. Channels are typed, and can only transfer values of a single type. Therefore, we use one-shot channels, and encode sessions following \textcite{dardha2017,scalas2016,padovani2017}: each \lstinline{send} creates a new channel for the continuation of the session, and sends that along with the value, and each \lstinline{recv} receives a value together with a channel on which to continue the session. Fortunately, the channels in \lstinline{std::sync::mpsc} are highly optimised for one-shot usage.\footnote{\url{https://doc.rust-lang.org/src/std/sync/mpsc/mod.rs.html\#147-149}}

\subsection{What If I Want to Be Evil?}\label{sec:limitations}
There are three ways in which you can make code written using Rusty Variation deadlock. The first is by looping, which we discussed earlier. The other two are explicit, and surprisingly, \lstinline{unsafe} is not one of them:
